### 2020-05-17 paper.tex
Edit

IDEs as well as the developer tools of modern web browsers provide battle-tested debugger utilities \footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} to inspect frontend application code at runtime. With an imperative programming model in mind, they let the user jump from program statement to statement, stop execution upon hitting a breakpoint while having a closer look at the values currently assigned to variables. By the time they are even able to preserve call stacks for asynchronously executed code which was a hassle before \footnote{\url{https://developers.google.com/web/tools/chrome-devtools/javascript/reference\#call-stack}}.

@@

We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex. Three subjects even claimed using specialized logging functionalities provided through rxjs-spy. This library allows to ``tag'' specific points in the \mintinline{TypeScript}{Observable} graph with a probe. In addition to report values traversing the graph, each probe notifies about events when an observer subscribes or unsubscribes to them or when the source \mintinline{TypeScript}{Observable} completes \cite{rxjslifecycle}. As the subjects stated, the knowledge about these additional events helps in situations where they need to find problems within a graph composing multiple asynchronous data sources on multiple levels. Without it, they were blind to what actually happens within the system.

@@

Salvaneschi et al. did focus their research heavily on RP during the mid 2010s. Beside the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well. In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al. elaborate on the necessity of RP debugging tools which can understand the specific language abstractions. With the plugin \emph{Reactive Inspector} for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala. One of its main features is the visualization of the internal dependency graph. The new class of  breakpoints, the ``reactive breakpoints'', builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather then on a specific node in the graph. The execution of the program can be interrupted e.g. once a node retains a specific output value or an event of interest takes place in the runtime. Along with the RP specific debugging features, \emph{Reactive Inspector} implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.

@@

\emph{rxfiddle} proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}. The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g. when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s. An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time. They are considered valuable for documentation \cite{marblediagrams} as well as for testing \cite{marbletesting}} for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.

@@

Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier. \emph{rxjs-spy}\cite{rxjsspy} wants to improve on this by introducing the \mintinline{TypeScript}{tag} operator. Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are created right at the moment the \mintinline{TypeScript}{Observable} itself is defined. A tag as shown in Listing~\ref{lst:rxjs-spy-tag} does not influence runtime behavior at first.

@@

Later, \emph{rxjs-spy} can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all. It is worth to mention that manual log traces, without additional effort, only show values flowing by. The additional events logged by \emph{rxjs-spy} proofed being invaluable in our own experience multiple times.

The library provides further features to interact with tagged \mintinline{TypeScript}{Observable}s through the debuggers console.

@@

\subsection{Rx Vizualizer}

The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser. It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams.

>>>>

IDEs as well as the developer tools of modern web browsers provide battle-tested debugger utilities\footnote{\url{https://developer.mozilla.org/en-US/docs/Tools/Debugger}} to inspect frontend application code at runtime. With an imperative programming model in mind, they let the user jump from program statement to statement, stop execution upon hitting a breakpoint while having a closer look at the values currently assigned to variables. By the time they are even able to preserve call stacks for asynchronously executed code which was a hassle before\footnote{\url{https://developers.google.com/web/tools/chrome-devtools/javascript/reference\#call-stack}}.

@@

We could verify previously gathered interview data with the results from the new reports: The subjects tend to manually add log statements to their code in order to understand its behavior, especially when things get more complex. Three subjects even claimed using specialized logging functionalities provided by rxjs-spy\cite{rxjsspy}. As the subjects stated, the knowledge about additional events made visible helped in situations where they needed to find problems within a graph composing multiple asynchronous data sources on multiple levels. Without it, they would have been blind to what actually happened within the system.

@@

Salvaneschi et al. did focus their research heavily on RP during the mid 2010s. Beside the creation of REScala in 2014 \cite{10.1145/2577080.2577083}, an RP DSL and runtime for the Scala programming language, they looked into the debugging habits of software engineers as well. In the paper ``Debugging for Reactive Programming'' \cite{10.1145/2884781.2884815}, Salvaneschi et al. elaborate on the necessity of RP debugging tools which can understand the specific language abstractions. With the plugin Reactive Inspector for the Eclipse IDE, they provide a debugger which understands the RP language abstractions of REScala. One of its main features is the visualization of the internal dependency graph. The new class of  breakpoints, the ``reactive breakpoints'', builds upon the visualized graph as well: They do not depend on a specific line of code or statement rather then on a specific node in the graph. The execution of the program can be interrupted e.g. once a node retains a specific output value or an event of interest takes place in the runtime. Along with the RP specific debugging features, Reactive Inspector implements omniscient debugging concepts as well: Time travel allows to record and replay runtime behavior at a later time without reexecution.

@@

rxfiddle proposed by Banken et al. is the first reference implementation of their RP debugger architecture for the ReactiveX API specification described in their paper ``Debugging Data Flows in Reactive Programs'' \cite{10.1145/3180155.3180156}. The design defines two independent parts: The (i) \emph{Host instrumentation} instruments a ReactiveX implementation to emit events at runtime (e.g. when an \mintinline{TypeScript}{Observable} is created or subscribed) which get collected. The (ii) \emph{Visualizer} component interprets these events and presents them along two dimensions: The StoryFlow graph \cite{YWu2013a} shows when an \mintinline{TypeScript}{Observable} is created and how it interacts with other \mintinline{TypeScript}{Observable}s. An additional marble diagram\footnote{Marble diagrams are used in the ReactiveX community to describe the behavior of an \mintinline{TypeScript}{Observable} over its life time. They are considered valuable for documentation \cite{marblediagrams} as well as for testing \cite{marbletesting}} for every \mintinline{TypeScript}{Observable} is used to display the values it emits over time.

@@

Software engineers tend to augment their code with additional log trace statements in the situation where they have to track down a problem as we saw earlier. rxjs-spy\cite{rxjsspy} intends to improve on this by introducing the \mintinline{TypeScript}{tag} operator. Instead of adding log traces at debug-time, ``tagged'' \mintinline{TypeScript}{Observable}s are created right at the moment the \mintinline{TypeScript}{Observable} itself is defined. A tag as shown in Listing~\ref{lst:rxjs-spy-tag} does not influence runtime behavior at first.

@@

Later, rxjs-spy can be advised to log events such as subscribe, emit, error, complete and unsubscribe of matching tags directly to the console without touching any \mintinline{TypeScript}{Observable} code at all. It is worth to mention that manual log traces, without additional effort, only show values flowing by. The additional events logged by rxjs-spy were noted invaluable in the interviews presented earlier as well as in our own experience.

@@

\subsection{Rx Visualizer}

The \emph{Rx Visualizer}\cite{rxviz} is a utility available online in the browser. It is an ``animated playground for Rx Observables''\cite{rxviz} and allows to visualize snippets of RxJS code using marble diagrams. Diagrams are generated over a given, fixed time and can be saved as SVG\footnote{\url{https://developer.mozilla.org/en-US/docs/Web/SVG}}.

The visualizer helps to understand the behavior of a specific part of an \mintinline{TypeScript}{Observable} stream in a sandboxed environment.
